---
title: "Diagnostics for 2SLS Regression"
author: "John Fox"
date: "2019-07-26"
bibliography: Diagnostics-for-2SLS-Regression.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics-for-2SLS-Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The **lm2sls** package extends a variety of standard numeric and graphical regression  diagnostics to linear models fit by two-stage least-squares (2SLS) regression, a commonly employed method of instrumental-variables estimation for potentially overidentified structural equations in which there are endogenous regressors. The `fit2sls()` function in the package computes the 2SLS estimator employing a low-level interface not generally intended for direct use, and returns a list containing quantities that faciliate the computation of various diagnostics. The `lm2sls()` function provides a user-friendly formula-based interface to `fit2sls()`.

`lm2sls()` is broadly similar in its usage to the `tsls()` function in the **sem** package and the `ivreg()` function in the **AER** package. Its raison-d'etre therefore is the provision of regression diagnostics. The subject of this vignette is the rationale for the various diagnostics and the use of functions in the **lm2sls** package to compute them, along with functions in other packages (specifically the base-R **stats** package and the **car** and **effects** packages) that work with the `"2sls"` objects produced by `lm2sls()`.

## Case-Deletion Diagnostics for 2SLS Regression

As far as I can tell, diagnostics for regression models fit by 2SLS is a relatively neglected topic, but it was addressed briefly by @Belsley-Kuh-Welsch-1980 [pp. 266--268].


```{r setup}
library(lm2sls)
```


## References
